Sequential rank agreement methods for comparison of lists

### 2015-08-08 latex/paper.Rnw
Fixing comments

Incomplete lists are also a common occurrence. They arise for example
in case of missing data (items), when comparing top $d$ list results
from publications, or when some methods only rank a subset of the
items. For example, penalized regression based on the LASSO provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the set of predictors whose coefficient has been
shrunk to zero. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list.

@@

Four classification methods are considered: Random Forest
\citep{breiman2001random} implemented in the R package
\texttt{randomForest} \citep{liaw2002classification}, logistic LASSO
\citep{tibshirani1996regression} and Ridge regression
\citep{segerstedt1992ordinary} both implemented in the R package
\texttt{glmnet} \citep{friedman2010regularization}, and Partial Least
Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls}
implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

@@

\subsection{Agreement of predictor rankings}

For each of the four methods, each of the 1000 models trained in on of
the 1000 sub-samples of the MALOVA data produces a ranking of the 5000
predictors according to their importance for discriminating between
the tumor types.

All four methods depend on a tuning parameter. We select these
separately for each sub-sample by a 20 times repeated 5-fold
cross-validation procedure. For Lasso and Ridge regression the tuning parameter
is the degree of penalization, for PLS-DA it is the number of components 
(the dimensionality of the subspace), and for Random Forest it is the number
 of predictors randomly sampled at each split.

The implementation of LASSO and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreement.
Note also that the LASSO models produce censored lists depending on
the value of the penalty parameter. Under the deviance optimization
criterion the median number of non-zero coefficients was 33 (range 18
to 52). In order to calculate the sequential rank agreement for the
LASSO we therefore applied our proposed Algorithm \ref{sra-algorithm}
with $B = 1000$ and $P=5000$.

For the Random Forest classifier the predictors are ranked according
to the Gini index, while for the logistic LASSO and Ridge regression
models we order the predictors by absolute magnitude of the estimated
regression coefficients. For the PLS-DA model the importance of the
predictors is based on a weighted sum of the absolute coefficients
where the weights are proportional to the reduction in the sums of
squares across the components.

The right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
120. At deeper list depths all agreement curves appeared stable.

@@

For most of the sequential rank agreement curves we see, as expected,
that they start low, indicating good agreement, followed by an
increase until they approximately become constant. This has the
interpretation that the agreement across the different sub-samples is
higher in the top as compared to the tail of the lists for all these
classification methods. The change-points where the curves become
approximately constant are the list depths where the ranks of the
remaining items become close to random.
%% ANDREAS: boer "close to random" ikke vaere ens for alle metoder?

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
%% ANDREAS: er det sidste ok?
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by artificially censoring all predictors with absolute
coefficient values below a given threshold. For the Ridge models tuned
with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The corresponding curve from Figure
\ref{fig:app1} (left panel) is shown for comparison.
%% ANDREAS: godkender du min aendring af de sidste 2 saetninger?
Thus, in our application the median number of censored predictors
across the 1000 sub-samples was 5 (range 1 to 11). Even though the
number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing. Also, the effect of
the censoring caused by the sparsity of the LASSO is clearly visible
in the agreement curves.
%% ANDREAS: man kan desvaerre ikke se hvor LASSO agreement forsvinder
%% (ved list-depth 5?) og om den er laver for list depth 1-5 end rigde
%% eller om den er kun laver for list depth 1-2. maaske kan du
%% beskrive dette i teksten?

\subsection{Agreement of individual risk predictions}
\label{sec:airp}

To assess the stability of the individual risk predictions we apply
the predictors from the DACOVA data set to each of the models. The
predicted probabilities are then ranked in decreasing order such the
patients with the highest risk of a malignant tumor appears on the top
of the list. Figure \ref{fig:app1} (left) shows the sequential rank
agreement separately for each method, based on 1000 risk predictions
obtained from the models trained in the same $1000$ random sub-samples
of the MALOVA data. For clarity of presentation we zoom in on the
agreement up to list depth 600. At deeper list depths all agreement
curves appeared stable.

Most curves start low and then increase indicating higher agreement
among high risk patients.
%% ANDREAS og CLAUS: er det ogsaa forvented eller skyldes det at vi kumulerer fra toppen?
%% agreement boer jo igen vaere hoejt bland low risk patienter ...
An exception is the risk prediction agreement for the LASSO tuned with
the AUC criterion which shows very low agreement among the high values
of the predicted risks. The reason is that optimizing the penalty
parameter with respect to the AUC criterion tends to favor a very high
penalty value causing only a single predictor to be selected in each
of the 1000 iterations. This results in a lack of generalizability to
the DACOVA data which gives rise to the higher disagreement in the
predicted risk probabilities. In the extreme case where the penalty
becomes so high that none of the predictors are selected by the LASSO,
the sequential rank agreement for the predicted probabilities becomes
undefined since all the ranks will be ties.

@@

In Figure \ref{fig:app1} it can further be seen that some of the
methods show better agreement with respect to the predicted
probabilities than for ranking the importance of the predictors and
vice versa. Ridge regression shows higher agreement across training
sets for the risk predictions than PLS-DA, and PLS-DA shows higher
agreement for predictor importance than Ridge regression. The Random
Forest method seems to perform equally well in both scenarios.
%% Andreas: jeg forstaar ikke den sidste saetning. kan man sammenligne de to rank agreements (left vs right panel)?
LASSO shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poor agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the LASSO penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable is selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.71$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
%% and for this data set it appears that PLS-DA is the overall best
%% method taking both agreement and discrimination ability into account.
%% Andreas: vi boer ogsaa se paa Brier score (calibration) inden vi konkluderer at PLS-DA er den bedste

@@

We note that the sequential rank agreement is still marred by problems
that generally apply to ranking of items and/or
individuals. Collinearity in particular can be a huge problem when
bootstrapping data or when comparing different analysis methods. For
example, marginal analyses where each item is analyzed separately will
assign similar ranks to two highly correlated predictors while methods
that provide a sparse solution such as the LASSO will just rank one of
the two predictors high while the other might have a very low rank.
Thus in such a scenario we would expect low agreement of the rankings
from LASSO and marginal analyses simply because of the way correlated
predictors are handled. This is not a shortcoming of the sequential
rank agreement but is a problem general to all ranked lists.

>>>>

Incomplete lists are also a common occurrence. They arise for example
in case of missing data (items), when comparing top $d$ list results
from publications, or when some methods only rank a subset of the
items. For example, penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the set of predictors whose coefficient has been
shrunk to zero. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list.

@@

Four classification methods are considered: Random Forest
\citep{breiman2001random} implemented in the R package
\texttt{randomForest} \citep{liaw2002classification}, logistic Lasso
\citep{tibshirani1996regression} and Ridge regression
\citep{segerstedt1992ordinary} both implemented in the R package
\texttt{glmnet} \citep{friedman2010regularization}, and Partial Least
Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls}
implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

@@

All four methods depend on a tuning parameter. We select these
separately for each sub-sample by a 20 times repeated 5-fold
cross-validation procedure. The tuning parameter for Lasso and Ridge regression 
is the degree of penalization, for PLS-DA it is the number of components 
(the dimensionality of the subspace), and for Random Forest it is the number
 of predictors randomly sampled at each split.
The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce censored lists depending on
the value of the penalty parameter.

\subsection{Agreement of predictor rankings}
For each of the four methods, each of the 1000 models trained in on of
the 1000 sub-samples of the MALOVA data produces a ranking of the 5000
predictors according to their importance for discriminating between
the tumor types. For the Random Forest classifier the predictors are ranked according
to the Gini index, while for the logistic Lasso and Ridge regression
models we order the predictors by absolute magnitude of the estimated
regression coefficients. For the PLS-DA model the importance of the
predictors is based on a weighted sum of the absolute coefficients
where the weights are proportional to the reduction in the sums of
squares across the components.

The right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
600. At deeper list depths all agreement curves become approximately constant.

@@

For most of the sequential rank agreement curves we see, as expected,
that they start low, indicating good agreement, followed by an
increase until they approximately become constant. This has the
interpretation that the agreement across the different sub-samples is
higher in the top as compared to the tail of the lists for all these
classification methods. The change-points where the curves become
approximately constant are the list depths where the ranks of the
remaining items become close to uniformly random.

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by artificially censoring all predictors with absolute
coefficient values below a given threshold. For the Ridge models tuned
with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The corresponding curve from Figure
\ref{fig:app1} (left panel) is shown for comparison.
%%%% CHECK UP: CENSOR ON POOLED DATA OR CENSOR BY INVIDIAL THRESHOLD

%%%%%% DET HER ER NOGET ROD. DER ER FORSVUNDET NOGET LASSO TEKST
%%%%%% GÅ TILBAGE OG FIND DET IGEN
Under the deviance optimization criterion the median number of non-zero 
coefficients was 33 (range 18 to 52). In order to calculate the sequential 
rank agreement for the Lasso we therefore applied our proposed Algorithm \ref{sra-algorithm}
with $B = 1000$ and $P=5000$.

Thus, in our application the median number of censored predictors
across the 1000 sub-samples was 5 (range 1 to 11). Even though the
number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing. Also, the effect of
the censoring caused by the sparsity of the Lasso is clearly visible
in the agreement curves.
\textbf{ANDREAS: man kan desvaerre ikke se hvor Lasso agreement forsvinder
  (ved list-depth 5?) og om den er laver for list depth 1-5 end rigde
  eller om den er kun laver for list depth 1-2. maaske kan du
  beskrive dette i teksten?}

\subsection{Agreement of individual risk predictions}
\label{sec:airp}
To assess the stability of the individual risk predictions we apply
the predictors from the DACOVA data set to each of the models. The
predicted probabilities are then ranked in decreasing order such that the
patients with the highest risk of a malignant tumor appears in the top
of the list. Figure \ref{fig:app1} (left) shows the sequential rank
agreement separately for each method, based on the 1000 risk predictions
obtained from the models trained in the same $1000$ random sub-samples
of the MALOVA data.

Most curves start low and then increase indicating higher agreement
among high risk patients.
\textbf{ANDREAS og CLAUS: er det ogsaa forvented eller skyldes det at vi kumulerer fra toppen? agreement boer jo igen vaere hoejt bland low risk patienter ... Claus indsætter måske noget her!}
An exception is the risk prediction agreement for the Lasso tuned with
the AUC criterion which shows very low agreement among the high values
of the predicted risks. The reason is that optimizing the penalty
parameter with respect to the AUC criterion tends to favor a very high
penalty value causing only a single predictor to be selected in each
of the 1000 iterations. This results in a lack of generalizability to
the DACOVA data which gives rise to the higher disagreement in the
predicted risk probabilities. In the extreme case where the penalty
becomes so high that none of the predictors are selected by the Lasso,
the sequential rank agreement for the predicted probabilities becomes
undefined since all the ranks will be ties.

@@

In Figure \ref{fig:app1} it can further be seen that some of the
methods show better agreement with respect to the predicted
probabilities than for ranking the importance of the predictors and
vice versa. Ridge regression shows higher agreement across training
sets for the risk predictions than PLS-DA, and PLS-DA shows higher
agreement for predictor importance than Ridge regression. The Random
Forest method seems to perform equally well in both scenarios.
\textbf{Andreas: jeg forstaar ikke den sidste saetning. kan man sammenligne de to rank agreements (left vs right panel)?}
Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poor agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the Lasso penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable is selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.71$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
\textbf{and for this data set it appears that PLS-DA is the overall best
  method taking both agreement and discrimination ability into account.
  Andreas: vi boer ogsaa se paa Brier score (calibration) inden vi konkluderer at PLS-DA er den bedste
}

@@

We note that the sequential rank agreement is still marred by problems
that generally apply to ranking of items and/or
individuals. Collinearity in particular can be a huge problem when
bootstrapping data or when comparing different analysis methods. For
example, marginal analyses where each item is analyzed separately will
assign similar ranks to two highly correlated predictors while methods
that provide a sparse solution such as the Lasso will just rank one of
the two predictors high while the other might have a very low rank.
Thus in such a scenario we would expect low agreement of the rankings
from Lasso and marginal analyses simply because of the way correlated
predictors are handled. This is not a shortcoming of the sequential
rank agreement but is a problem general to all ranked lists.
